---
title: "Triaging convergence diagnostics with diagnose_convergence()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Triaging convergence diagnostics with diagnose_convergence()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("cmdstanr", quietly = TRUE) &&
    tryCatch({ cmdstanr::cmdstan_path(); TRUE }, error = function(e) FALSE)
)
```

## Why tiered triage

In a `bilatr` panel fit, dyads vary enormously in how much data they
contribute: some have hundreds of observed periods, others just a
handful. A sparse dyad's own latent trajectory (`theta`) is *expected*
to have a wide, poorly-identified posterior -- that's not a bug, it's
what little data looks like. Reading `fit$summary()` or
`fit$diagnostic_summary()` directly makes it easy to either drown in
that expected noise, or (worse) miss a real problem in a global/shared
parameter because it's buried among thousands of noisy per-dyad-period
rows.

[diagnose_convergence()] addresses this by classifying every monitored
quantity into one of three tiers before applying Rhat/ESS thresholds:

* **Tier 1 -- global/shared parameters** (`alpha`, `mu_intercept`,
  `mu_theta0`, `sigma_theta0`, `mu_log_phi`, `sigma_log_phi`,
  `mu_log_noise`, `sigma_log_noise`, and `lp__`). These affect the whole
  model, so any issue here is always reported in full, never summarized
  away.
* **Tier 2 -- per-dyad hierarchical parameters** (`phi`, `process_noise`,
  `theta0`, and their non-centered/raw counterparts). These are screened
  against how much data (`n_dt`) each dyad has, flagging dyads whose
  diagnostics are worse than a smooth degradation-with-sparsity trend
  would predict -- i.e. genuine outliers, not just the sparsest dyads.
* **Tier 3 -- per-dyad-period latent states** (`theta`, `theta_raw`).
  These are the noisiest by design, so they're reported only as
  per-dyad aggregate statistics (e.g. minimum ESS along a dyad's path),
  never flooding the console with one row per dyad-period.

## A panel with very uneven sparsity

We simulate a small panel where dyad activity spans two orders of
magnitude, to make the tiering concrete:

```{r}
library(bilatr)
library(dplyr)

set.seed(42)
countries <- c("USA", "CHN", "RUS", "IRN", "PRK", "ISR")
dyad_pairs <- expand.grid(a1 = countries, a2 = countries, stringsAsFactors = FALSE) %>%
  filter(a1 != a2) %>%
  slice(1:8)

# very uneven event counts per dyad: some dyads get a handful of events,
# others hundreds.
n_events_per_dyad <- c(400, 350, 300, 250, 15, 10, 8, 5)
events <- purrr::map2_dfr(seq_len(nrow(dyad_pairs)), n_events_per_dyad, function(i, n) {
  dates <- format(
    sample(seq(as.Date("2015-01-01"), as.Date("2019-12-31"), by = "day"), n, replace = TRUE),
    "%Y%m%d"
  )
  tibble::tibble(
    Actor1CountryCode = dyad_pairs$a1[i],
    Actor2CountryCode = dyad_pairs$a2[i],
    SQLDATE = as.integer(dates),
    EventCode = sample(cameo_lookup$CAMEOEVENTCODE, n, replace = TRUE)
  )
})
events <- recode_cameo(events, code_col = "EventCode")

stan_data <- assemble_stan_data(
  events,
  years = 2015:2019,
  resolution = "yearly",
  grouping_var = "PentaClass",
  reference_category = 0,
  reference_hostile = 4,
  min_n_events = 1,
  chunk_size = 4
)
stan_data$D
```

```{r, results = "hide"}
fit <- fit_panel(
  stan_data,
  chains = 2,
  parallel_chains = 2,
  threads_per_chain = 2,
  iter_warmup = 300,
  iter_sampling = 300,
  seed = 1,
  opt_level = 1
)
```

As in the other vignettes, this run is deliberately short (and the event
data purely random, with no real signal for the model to recover) so
the vignette builds quickly. That means it's a poor fit in an absolute
sense -- don't be surprised to see Tier 1 flags below that a longer,
real run wouldn't show. That's still a useful illustration: on a real
fit, any Tier 1 flag is exactly the kind of thing this tool is meant to
surface loudly rather than let it get lost among per-dyad noise.

## Running the triage

`diagnose_convergence()` needs `n_dt`: the total observation count per
dyad, keyed by the same dyad index (1..D) used inside the Stan model.
Here we derive it directly from the `Y` array already assembled by
[assemble_stan_data()]:

```{r}
n_dt <- tibble::tibble(
  dyad_id = seq_len(stan_data$D),
  n_dt = apply(stan_data$Y, 1, sum)
)
n_dt

diag <- diagnose_convergence(fit, n_dt = n_dt)
diag
```

The printed report separates concerns by tier: Tier 1 issues (if any)
are always shown in full; Tier 2 shows a compact per-dyad table sorted
so any dyad flagged as "worse than expected for its `n_dt`" appears
first; Tier 3 collapses to one-line aggregate statistics.

## Working with the pieces directly

`diag` is a plain list (class `bilatr_diagnostics`) with `tier1`,
`tier2`, `tier3`, and `summary` elements, so you can filter or
join further as needed:

```{r}
# any global/shared parameter that failed to converge:
dplyr::filter(diag$tier1, flagged)

# dyads whose per-dyad parameters look worse than their n_dt would predict:
dplyr::filter(diag$tier2, worse_than_expected)

# headline counts:
diag$summary
```

## Computing only some tiers

`tiers` (default `1:3`) restricts which tier(s) are computed at all --
useful since Tier 3 typically dwarfs the other two in the number of
monitored quantities (one per dyad-period), so computing Rhat/ESS for
it is the most expensive part of a call. A quick check of just the
global/shared parameters, with no `n_dt` needed at all since Tier 1 has
no per-dyad structure to join against it, looks like:

```{r}
diagnose_convergence(fit, tiers = 1)
```

`diag$tier2`/`diag$tier3` are `NULL` (not empty tibbles) for any tier
left out of `tiers`, so `is.null(diag$tier3)` distinguishes "not
computed" from "computed, nothing to report".

## Tuning thresholds

`rhat_threshold` (default `1.01`) and `ess_threshold` (default `400`,
applied to both `ess_bulk` and `ess_tail`) follow Vehtari et al. (2021)
and can be tightened or loosened:

```{r, eval = FALSE}
diagnose_convergence(fit, n_dt = n_dt, rhat_threshold = 1.05, ess_threshold = 100)
```

## On the "worse than expected" flag

The Tier 2/3 outlier flag is deliberately simple: dyads are grouped
into `n_dt` quantile brackets, and a dyad is flagged when its ESS falls
more than 1.5 times the interquartile range below its own bracket's
median (the standard Tukey outlier rule). This means a very sparse
dyad with a wide, low-ESS posterior is *not* flagged as long as its
peers at similar sparsity look about the same -- only a dyad that
stands out *relative to equally-sparse peers* is surfaced, which is
the actual failure mode worth investigating (e.g. a data or
identification issue specific to that dyad, rather than sparsity
alone).
