---
title: "Single-dyad time series with bilatr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-dyad time series with bilatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("cmdstanr", quietly = TRUE) &&
    tryCatch({ cmdstanr::cmdstan_path(); TRUE }, error = function(e) FALSE)
)
```

This vignette walks through fitting `bilatr`'s dyadic IRT model to a
**single dyad's** event time series: data prep, fitting, posterior
extraction, and basic convergence diagnostics. For fitting across many
dyads at once with hierarchical pooling, see `vignette("panel_model")`.

## 1. Data prep

In a real workflow you'd start from raw GDELT or ICEWS event data:

```{r, eval = FALSE}
library(bilatr)

# one CAMEO-coded event file
events <- extract_all_relevant_gdelt("data/gdelt_raw/20200101.zip")
events <- recode_cameo(events)
```

For this vignette we simulate a small, self-contained event set for one
dyad so the example runs without a network connection or a local GDELT
archive:

```{r}
library(bilatr)
library(dplyr)

set.seed(42)
n_events <- 400
dates <- format(
  sample(seq(as.Date("2015-01-01"), as.Date("2019-12-31"), by = "day"), n_events, replace = TRUE),
  "%Y%m%d"
)
events <- tibble::tibble(
  Actor1CountryCode = "USA",
  Actor2CountryCode = "CHN",
  SQLDATE = as.integer(dates),
  EventCode = sample(cameo_lookup$CAMEOEVENTCODE, n_events, replace = TRUE)
)
events <- recode_cameo(events, code_col = "EventCode")
```

`recode_cameo()` attaches `QuadClass`, `PentaClass`, and
`PentaClass_modified` from the package's built-in [cameo_lookup] table
(no external CAMEO reference package required).

## 2. Assemble Stan data

[assemble_stan_data()] aggregates events to dyad-period action-class
counts and packages them for the model. `reference_category` and
`reference_hostile` anchor the model's identification: the first should
be a low-conflict/neutral action class, the last should be the most
conflictual. Here we use PentaClass 0 (verbal cooperation) and 4
(the most severe conflict class).

```{r}
stan_data <- assemble_stan_data(
  events,
  years = 2015:2019,
  resolution = "yearly",
  grouping_var = "PentaClass",
  reference_category = 0,
  reference_hostile = 4,
  min_n_events = 1
)
stan_data$D # a single dyad
stan_data$T # number of years
stan_data$A # number of action classes
```

Leaving `dyad_weight`/`period_weight`/`action_weight` unset (as above)
defaults them to vectors of 1s, i.e. the model's base, unweighted form.

## 3. Fit

[fit_dyad_ts()] checks that `stan_data$D == 1` and fits the model with
`threads_per_chain = 1` by default (reduce_sum has nothing to chunk
across when there's only one dyad). The example below uses a short
warmup/sampling run so the vignette builds quickly; real analyses should
use `iter_warmup`/`iter_sampling` in the low thousands and check the
diagnostics below before trusting the result.

```{r, results = "hide"}
fit <- fit_dyad_ts(
  stan_data,
  chains = 2,
  parallel_chains = 2,
  iter_warmup = 300,
  iter_sampling = 300,
  seed = 1,
  opt_level = 1
)
```

## 4. Extract

[extract_theta()] reattaches human-readable dyad identifiers to the
posterior summary of the latent conflict trajectory:

```{r}
theta <- extract_theta(fit, stan_data)
theta
```

[extract_alpha()] and [extract_mu_intercept()] extract the action-type
discrimination and intercept parameters, optionally labeled with the
event-class values used to build `stan_data`:

```{r}
extract_alpha(fit, event_classes = attr(stan_data, "event_classes"))
extract_mu_intercept(fit, event_classes = attr(stan_data, "event_classes"))
```

Note `alpha[1] = 1` and `mu_intercept[1] = 0` exactly: these are the
model's fixed identification anchors, not estimated quantities.

## 5. Diagnose

`cmdstanr`'s default posterior summary already includes R-hat and both
bulk and tail effective sample size; check these before drawing
conclusions from `theta`:

```{r}
fit$summary(variables = c("alpha", "mu_intercept", "phi", "process_noise"))
```

Look for `rhat` close to 1 (below ~1.01) and `ess_bulk`/`ess_tail` in the
hundreds or more per chain. `fit$diagnostic_summary()` reports any
divergent transitions or tree-depth saturation, which for a real fit
should be zero or reported here:

```{r}
fit$diagnostic_summary()
```

If ESS is low or R-hat is elevated, the usual fixes are more sampling
iterations, tighter priors informed by domain knowledge, or (for
single-dyad fits especially, since there's no cross-dyad pooling to help
identify `theta0`/`phi`) more data.
