---
title: "Panel estimation across multiple dyads with bilatr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panel estimation across multiple dyads with bilatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("cmdstanr", quietly = TRUE) &&
    tryCatch({ cmdstanr::cmdstan_path(); TRUE }, error = function(e) FALSE)
)
```

This vignette walks through fitting `bilatr`'s dyadic IRT model across
**multiple dyads at once**, with hierarchical partial pooling on process
noise, the dispersion parameter `phi`, and the initial latent state
`theta0`. For a single dyad, see `vignette("dyad-time-series")`; the
data-prep and extraction steps are the same, only the fitting call and
the presence of pooling differ.

## 1. Data prep

As in the single-dyad case, a real workflow starts from raw event data
recoded via [recode_cameo()]. Here we simulate a small panel of six
dyads so the example is self-contained:

```{r}
library(bilatr)
library(dplyr)

set.seed(42)
countries <- c("USA", "CHN", "RUS", "IRN")
dyad_pairs <- expand.grid(a1 = countries, a2 = countries, stringsAsFactors = FALSE) %>%
  filter(a1 != a2)

n_events <- 3000
dates <- format(
  sample(seq(as.Date("2015-01-01"), as.Date("2019-12-31"), by = "day"), n_events, replace = TRUE),
  "%Y%m%d"
)
pair_idx <- sample(nrow(dyad_pairs), n_events, replace = TRUE)
events <- tibble::tibble(
  Actor1CountryCode = dyad_pairs$a1[pair_idx],
  Actor2CountryCode = dyad_pairs$a2[pair_idx],
  SQLDATE = as.integer(dates),
  EventCode = sample(cameo_lookup$CAMEOEVENTCODE, n_events, replace = TRUE)
)
events <- recode_cameo(events, code_col = "EventCode")
```

## 2. Assemble Stan data

The call is identical in shape to the single-dyad case; `D` will simply
be greater than 1 because the input spans multiple dyads. Panel-sized
data is where `reduce_sum` chunking matters: `chunk_size` sets how many
dyads' worth of likelihood evaluation each `reduce_sum` task covers.
This project's own benchmarking found `chunk_size = 600` with 16 threads
Pareto-optimal for panel-sized data; for a small illustrative panel like
this one, a small `chunk_size` (or even 1) is fine.

```{r}
stan_data <- assemble_stan_data(
  events,
  years = 2015:2019,
  resolution = "yearly",
  grouping_var = "PentaClass",
  reference_category = 0,
  reference_hostile = 4,
  min_n_events = 1,
  chunk_size = 4
)
stan_data$D # multiple dyads
```

Setting `weighted = "all"` instead would additionally downweight
high-volume dyads/periods and upweight rare action classes, useful when
a handful of very active dyads would otherwise dominate the likelihood.
`weighted` also accepts single components or `"-"`-separated
combinations (e.g. `"dyad"`, `"dyad-period"`) if you only want some of
the three reweighting schemes computed, leaving the others at 1s:

```{r, eval = FALSE}
stan_data_weighted <- assemble_stan_data(
  events,
  years = 2015:2019,
  resolution = "yearly",
  grouping_var = "PentaClass",
  reference_category = 0,
  reference_hostile = 4,
  weighted = "all"
)

# only downweight high-volume dyads, leave period/action weights at 1s
stan_data_dyad_weighted <- assemble_stan_data(
  events,
  years = 2015:2019,
  resolution = "yearly",
  grouping_var = "PentaClass",
  reference_category = 0,
  reference_hostile = 4,
  weighted = "dyad"
)
```

## 3. Fit

[fit_panel()] checks that `stan_data$D >= 2` and defaults
`threads_per_chain = 16` (tune this and `chunk_size` together for your
hardware; see [fit_panel()]'s documentation). As in the single-dyad
vignette, this example uses a short run to build quickly.

```{r, results = "hide"}
fit <- fit_panel(
  stan_data,
  chains = 2,
  parallel_chains = 2,
  threads_per_chain = 2,
  iter_warmup = 300,
  iter_sampling = 300,
  seed = 1,
  opt_level = 1
)
```

## 4. Extract

The extraction functions work identically to the single-dyad case, now
returning one trajectory per dyad:

```{r}
theta <- extract_theta(fit, stan_data)
theta %>% count(dyad) # one row per dyad-period, D dyads

extract_alpha(fit, event_classes = attr(stan_data, "event_classes"))
```

The hierarchical pooling shows up in the dyad-level parameters, e.g.
`phi` (one value per dyad, pooled toward `mu_log_phi`/`sigma_log_phi`)
and `theta0` (pooled toward `mu_theta0`/`sigma_theta0`):

```{r}
fit$summary(variables = c("phi", "theta0", "mu_log_phi", "sigma_log_phi", "mu_theta0", "sigma_theta0"))
```

## 5. Diagnose

Convergence diagnostics work the same way as for a single dyad, but with
a panel it's worth checking whether any *specific* dyads are poorly
identified (e.g. very low event counts) rather than only looking at
global summaries:

```{r}
fit$diagnostic_summary()

theta %>%
  group_by(dyad) %>%
  summarise(mean_ess_proxy = mean(`95%` - `5%`)) %>%
  arrange(desc(mean_ess_proxy))
```

Dyads with unusually wide posterior intervals relative to the rest of
the panel are the ones most likely to need more data, a higher
`min_n_events` cutoff, or closer inspection of their raw event counts
before trusting their `theta` trajectory.
